\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-05-29 misc/JSS article/final/article.Rnw
Final article update.

% Hypthens
\hyphenation{mod-i-fi-ca-tions ma-nip-u-la-tion trans-for-ma-tion per-for-mance de-vel-op-ment
             ef-fi-cient in-tro-duces ca-pa-bil-i-ties gen-er-al-iz-ing en-abling pro-gram-ming
             com-pu-ta-tions eval-u-at-ed fa-cil-i-tate im-ple-men-ta-tions re-place-ment
             sub-set-ting de-com-po-si-tion al-go-rith-mic en-hanced mul-ti-ple coun-tries
             com-pu-ta-tion-al par-tic-u-lar-ly}
\usepackage[english]{babel}
% \sloppy
% \usepackage{microtype}

@@

# This is the benchmark function (janitor 2.2.0 is used)
bmark <- function(...) {
  bench::mark(..., min_time = 0.01, check = FALSE) |> janitor::clean_names() |>
  fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
  fmutate(expression = names(expression)) |> dapply(as.character) |> qDF()
}

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions, such as vectorized and verbose hash-joins or fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is powered by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifications to the namespace to mask/replace equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline

@@

\pkg{collapse} relates to and integrates key elements from these projects. It offers \pkg{tidyverse}-like data manipulation at the speed and stability of \pkg{data.table} for any data frame-like object. It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest}, and \pkg{xts}. It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}. Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and \proglang{C}-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal. It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}. It provides novel recursive tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr}, and \pkg{rrapply}. Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes, requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specialized statistics. \newline

@@

The package has a built-in structured \href{https://sebkrantz.github.io/collapse/reference/index.html}{documentation} facilitating its use. This documentation includes a central \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and supplementary topic pages which briefly describe related functionality. The names of these extra pages are collected in a global macro \code{.COLLAPSE\_TOPICS} and can be called directly with \code{help()}:

@@

This article does not fully present \pkg{collapse}, but the following sections introduce its key components, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} $+$ \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.

@@

fmean(num_vars(iris), g = iris$Species, nthreads = 4)

@@

Having grouping and data transformation functionality directly built into generic statistical functions facilitates and speeds up many common operations. It also avoids the need to convert atomic objects to data frames for grouped aggregations or transformations. The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://sebkrantz.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:

@@

With the exception of \code{pwcor(Dlog(LIFEEX), Dlog(PCGDP))}, which is evaluated for every country, all other expressions in \fct{fsummarise} and also the final \fct{collap} call are fully vectorized, i.e., \emph{Fast Statistical and Transformation Functions} are evaluated only once with grouping information passed to their \code{g} arguments. Using subtraction by reference in the range is more memory efficient as the executed expression is equivalent to \code{fmax(LIFFEX, country) \%-=\% fmin(LIFFEX, country)}. \fct{across} here directly invokes \fct{flast.data.frame} on the subset of columns.
\emph{Fast Statistical Functions} also have a method for grouped data, hence \fct{fsummarise} is not always needed. Below, I average the data with population weights.\footnote{Functions like \fct{fmean}, when called on a grouped data frame with weights, by default (\code{keep.w = TRUE}) sum the weights column and retain it after the grouping columns to permit further weighted operations. \vspace{-10mm}}

@@

\pkg{collapse} thus offers several ways to reach the same outcome. This may be confusing at first, but all options can be distinguished in terms of efficiency. It is thus helpful to be mindful of computer resources and attempt to craft minimalistic solutions to complex problems. For example, I recently combined multiple spatial datasets on points of interest (POIs) and needed to deduplicate them. I decided to keep the richest source for each location and POI type/category. After creating comparable POI confidence, location, and type indicators, my deduplication expression ended up being a single line of the form \code{fsubset(data, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (richest) source by location and type. This is efficient because it avoids materializing intermediate datasets and relegates all computations to \fct{fmode}---\emph{Fast Statistical Functions} internal grouped logic is highly optimized.
%
\subsection{Vectorizations for advanced tasks} \label{ssec:vfat}
%
\fct{fsummarise} and \fct{fmutate} follow an eager vectorization approach, such that using any \emph{Fast Statistical Function} in an expression causes the entire expression to be vectorized (i.e., evaluated only once with \emph{Fast Statistical Function} returning grouped output). This only applies to visible expressions, \pkg{collapse} cannot read the contents of custom functions, thus such functions are always evaluated using split-apply-combine logic. This eager vectorization approach enables efficient grouped calculations of more complex statistics. Below, I forecast the population of each region via linear regression (\code{POP ~ year}) in a fully vectorized way.
%
<<glinreg>>=
wlddev |> collap(POP ~ region + year, FUN = fsum) |>
  fmutate(POP = POP / 1e6) |> fgroup_by(region) |>
  fmutate(dmy = fmean(year, TRA = "-")) |>
  fsummarise(beta = fsum(POP, dmy) %/=% fsum(dmy, dmy),
             POP20 = flast(POP)) |>
  fmutate(POP21 = POP20 + beta, POP22 = POP21 + beta,
          POP23 = POP22 + beta, POP24 = POP23 + beta)
@
%
When \fct{fsummarise} evaluates an expression involving \emph{Fast Statistical Functions}, it sets their \code{g} argument with a grouping (\class{GRP}) object that is directly handed to \proglang{C/C++}, and also sets \code{use.g.names = FALSE}. Hence, weights (\code{w}) becomes the second positional argument. Similarly, \fct{fmutate} sets \code{g} and \code{TRA = "fill"}, which can be overwritten by the user (here with \code{TRA = "-"}). The expression \code{fsum(x, dmy) \%/=\% fsum(dmy, dmy)} amounts to \code{cov(x, y)/var(y)}, but is vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation. In a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{blog post}, I forecasted high-resolution population estimates for South Africa like this. Using 1 $km^2$ \href{https://www.worldpop.org/}{WorldPop} data available for years 2014-2020, I ran 1.6 million cell-regressions and obtained 2 forecasts for 2021 and 2022 in less than 0.3 seconds on my M1 Mac. Another neat example from the community, shared by Andrew Ghazi in a \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post}, vectorizes an expression to compute the $p$~value, \code{2 * pt(abs(fmean(x) * sqrt(6) / fsd(x)), 5, lower.tail = FALSE)}, across 300,000 groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}. \newline
% The eager vectorization approach of \pkg{collapse} here replaces \code{fmean(x)} and \code{fsd(x)} by their grouped versions and evaluates the entire expression once rather than 300k times as in \pkg{dplyr}. \newline

@@

Weighted quantiles have a sub-column level parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or the \code{nthreads} arguments to \fct{fnth}/\fct{fmedian}/\fct{fmode} (default 1).} and, as shown above, can harness an (optional) optimization utilizing an overall ordering vector (combining groups and the data column) to avoid repeated sorting of the same elements within groups. % \newline

@@

The lower-level API is useful for package development and standard-evaluation programming, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}. Users should note that \pkg{collapse} does not provide metaprogramming capabilities in its non-standard evaluation functions---such as \href{https://rlang.r-lib.org/reference/topic-quosure.html}{quosures} or \href{https://dplyr.tidyverse.org/articles/programming.html}{indirection} familiar to \pkg{tidyverse} users. Instead, it has standard-evaluation equivalents to some of these functions which typically end with a \code{v} for 'variables', such as \fct{collapv}, \fct{fslicev}, \fct{ftransformv}, etc. In other cases, including those typically handled by \fct{fsummarise} or \fct{fmutate}, users need to use the lower-level API for programming, or resort to \fct{substitute} and friends. The main reason for not providing higher-level metaprogramming capabilities is to keep all functions as simple as possible. It also compels users to think deeply about their programs and devise more efficient solutions. \newline

These metaprogramming restrictions are vindicated by internally complex data manipulation functions producing greater overheads that can noticeably add up in larger scripts:\footnote{\fct{bmark} is a slim wrapper around \fct{bench::mark}. See the Computational details section.}

@@

 bmark(collapse = collapse::fsummarise(mtcars, sum = sum(mpg)),
       dplyr = dplyr::summarise(mtcars, sum = sum(mpg)))

@@

Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices. For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 44.7 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}

@@

Time series functions such as \fct{fgrowth} to compute growth rates are S3 generic and can be applied to most time series classes. In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation. If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a plain numeric vector representing calendar years. Users need to manually call \code{timeid()} on plain numeric vectors with decimals to yield an appropriate integer representation. If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id. For the GCD approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}.} But first, a basic example:

@@

fgrowth(airmiles) |> round(2)
To illustrate the full capabilities of these time series functions, I generate a sector-level trade dataset of export values (v) by country (c), sector (s), and year (y). Like many detailed trade datasets, it is unbalanced---not all sectors/products are exported by country c in all years.
<<exports>>=
exports <- expand.grid(y = 2001:2010, c = paste0("c", 1:10),
  s = paste0("s", 1:10)) |> fmutate(v = abs(rnorm(1e3))) |>
  colorder(c, s) |> fsubset(-sample.int(1e3, 500))
The following extracts one country-sector series from the \code{exports} dataset. It is irregular, missing years 2003 and 2006.\footnote{\code{\%=\%} is an infix operator for the \fct{massign} function in \pkg{collapse} which is a multivariate version of \fct{assign}.} Indexation using the \code{t} argument then allows for correct (time-aware) computations in this context without the need to 'expand' the data/fill gaps.
<<ts_example_2>>=
.c(y, v) %=% fsubset(exports, c == "c1" & s == "s7", -c, -s)
print(y)
fgrowth(v, t = y) |> round(2)
fgrowth(v, -1:3, t = y) |> head(4)

@@

@@

These functions and operators are also integrated with \fct{fgroup\_by} and \fct{fmutate} for vectorized computations. But, as mentioned earlier, ad-hoc grouping is always more efficient.

@@

to the creation of new classes \class{indexes\_series} and \class{indexed\_frame} which inherit from the former. Any data frame-like object can become an \class{indexed\_frame} and behave as usual for other operations. The technical implementation of these classes is described in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{documentation}. Their basic syntax is:

@@

Indexation can be undone using \code{unindex()} and redone with \code{reindex()} and a suitable \class{index\_df}. \class{indexes\_series} can be atomic vectors or matrices (including objects such as \class{ts} or \class{xts}) and can be created directly using \code{reindex()}.

@@

bmark(base_int = unique(g_int), collapse_int = funique(g_int))

>>>>

% Hypthens
\hyphenation{mod-i-fi-ca-tions ma-nip-u-la-tion trans-for-ma-tion per-for-mance de-vel-op-ment
             ef-fi-cient in-tro-duces ca-pa-bil-i-ties gen-er-al-iz-ing en-abling pro-gram-ming
             com-pu-ta-tions eval-u-at-ed fa-cil-i-tate im-ple-men-ta-tions re-place-ment
             sub-set-ting de-com-po-si-tion al-go-rith-mic en-hanced mul-ti-ple coun-tries
             com-pu-ta-tion-al par-tic-u-lar-ly com-po-nents}
\usepackage[english]{babel}
% \sloppy
% \usepackage{microtype}

@@

# This is the benchmark function (janitor 2.2.0 is used)
bmark <- function(...) {
  bench::mark(..., min_time = 3, check = FALSE) |> janitor::clean_names() |>
  fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
  fmutate(expression = names(expression)) |> dapply(as.character) |> qDF()
}

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions, including vectorized and verbose hash-joins and fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is powered by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifications of the namespace to mask/replace equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline

@@

\pkg{collapse} relates to and integrates key elements from these projects. It offers \pkg{tidyverse}-like data manipulation at the speed and stability of \pkg{data.table} for any data frame-like object. It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest}, and \pkg{xts}. It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}. Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and \proglang{C}-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal. It also supports variable labels and intelligently preserves the attributes of all objects, complementing \pkg{labelled}. It provides novel recursive tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr}, and \pkg{rrapply}. Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes, requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specialized statistics. \newline

@@

The package comes with a built-in structured \href{https://sebkrantz.github.io/collapse/reference/index.html}{documentation} facilitating its use. This includes a central \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and a set of supplementary topic pages which briefly summarize related functionality. The names of these extra pages are collected in a global macro \code{.COLLAPSE\_TOPICS} and can be called directly with \code{help()}:

@@

While this article cannot fully present \pkg{collapse}, the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.

@@

fmean(num_vars(iris), g = iris$Species)

@@

Having grouping and data transformation functionality directly built into generic statistical functions facilitates and speeds up many common operations. It notably avoids the need to convert atomic objects to data frames for grouped aggregations or transformations. The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://sebkrantz.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:

@@

With the exception of \code{pwcor(Dlog(LIFEEX), Dlog(PCGDP))}, which is evaluated for every country, all other expressions in \fct{fsummarise} and also the final \fct{collap} call are fully vectorized, i.e., \emph{Fast Statistical and Transformation Functions} are evaluated only once with grouping information passed to their \code{g} arguments. Using subtraction by reference in the range is more memory efficient as the executed expression is equivalent to \code{fmax(LIFFEX, country) \%-=\% fmin(LIFFEX, country)}. \fct{across} directly invokes \fct{flast.data.frame} on the subset of columns. \emph{Fast Statistical Functions} also have a method for grouped data, thus \fct{fsummarise} is not always needed. Below, I average recent data with population weights.\footnote{Functions like \fct{fmean}, when called on a grouped data frame with weights, by default (\code{keep.w = TRUE}) sum the weights column and retain it after the grouping columns to permit further weighted operations. \vspace{-10mm}}

@@

\pkg{collapse} thus offers several ways to reach the same outcome. This may be confusing at first, but all options can be distinguished in terms of efficiency. Hence, it is helpful to think about computer resources and attempt to craft minimalistic solutions to complex problems. For example, I recently combined multiple spatial datasets on points of interest (POIs) and needed to deduplicate them. I decided to keep the richest source for each location and POI type/category. After creating comparable POI confidence, location, and type indicators, my deduplication expression ended up being a single line of the form \code{fsubset(data, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (richest) source by location and type. This is efficient because it avoids materializing intermediate datasets and relegates all computations to \fct{fmode}---\emph{Fast Statistical Functions} call a highly optimized internal grouping algorithm.
%
\subsection{Vectorizations for advanced tasks} \label{ssec:vfat}
%
\fct{fsummarise} and \fct{fmutate} follow an eager vectorization approach, such that using any \emph{Fast Statistical Function} in an expression causes the entire expression to be vectorized (i.e., evaluated only once with \emph{Fast Statistical Function} returning grouped output). This only applies to visible expressions, \pkg{collapse} cannot read the contents of custom functions, thus such functions are always evaluated using split-apply-combine logic. This eager vectorization approach enables efficient grouped calculations of more complex statistics. Below, I forecast the population of each region via linear regression (\code{POP ~ year}) in a fully vectorized way.
%
<<glinreg>>=
wlddev |> collap(POP ~ region + year, FUN = fsum) |>
  fmutate(POP = POP / 1e6) |> fgroup_by(region) |>
  fmutate(dmy = fmean(year, TRA = "-")) |>
  fsummarise(beta = fsum(POP, dmy) %/=% fsum(dmy, dmy),
             POP20 = flast(POP)) |>
  fmutate(POP21 = POP20 + beta, POP22 = POP21 + beta,
          POP23 = POP22 + beta, POP24 = POP23 + beta)
@
%
When \fct{fsummarise} evaluates an expression involving \emph{Fast Statistical Functions}, it sets their \code{g} argument with a grouping (\class{GRP}) object that is directly handed to \proglang{C/C++}, and also sets \code{use.g.names = FALSE}. Hence, weights (\code{w}) becomes the second positional argument. Similarly, \fct{fmutate} sets \code{g} and \code{TRA = "fill"}, which can be overwritten by the user (here with \code{TRA = "-"}). The expression \code{fsum(x, dmy) \%/=\% fsum(dmy, dmy)} amounts to \code{cov(x, y)/var(y)}, but is vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) internally and division by reference (\code{\%/=\%}) to avoid an additional allocation. In a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{blog post}, I forecasted high-resolution population estimates for South Africa like this. Using 1 $km^2$ \href{https://www.worldpop.org/}{WorldPop} data available for years 2014-2020, I ran 1.6 million cell-regressions and obtained 2 forecasts for 2021 and 2022 in less than 0.3 seconds on my M1 Mac. Another neat example from the community, shared by Andrew Ghazi in a \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post}, vectorizes an expression to compute the $p$~value, \code{2 * pt(abs(fmean(x) * sqrt(6) / fsd(x)), 5, lower.tail = FALSE)}, across 300,000 groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}. \newline
% The eager vectorization approach of \pkg{collapse} here replaces \code{fmean(x)} and \code{fsd(x)} by their grouped versions and evaluates the entire expression once rather than 300k times as in \pkg{dplyr}. \newline

@@

Weighted quantiles have a sub-column level parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or set the \code{nthreads} arguments of \fct{fnth}/\fct{fmedian} (default 1).} and, as shown above, can also harness an (optional) optimization via an overall ordering vector---combining groups with the data column to avoid repeated sorting of the same elements by different functions. % \newline

@@

The lower-level API is useful for package development and standard-evaluation programming, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}. Users should note that \pkg{collapse} does not provide metaprogramming capabilities in its non-standard evaluation functions---such as \href{https://rlang.r-lib.org/reference/topic-quosure.html}{quosures} or \href{https://dplyr.tidyverse.org/articles/programming.html}{indirection} familiar to \pkg{tidyverse} users. Instead, it has standard-evaluation equivalents to some of these functions which typically end with a \code{v} for 'variables', such as \fct{collapv}, \fct{fslicev}, \fct{ftransformv}, etc. In other cases, including those typically handled by \fct{fsummarise} or \fct{fmutate}, users need to use the lower-level API for programming, or resort to \fct{substitute} and friends. The main reason for not providing higher-level metaprogramming capabilities is to keep all functions as simple as possible. It also compels users to think more deeply about their programs and devise more efficient solutions. \newline

As the small benchmark below illustrates, \pkg{dplyr}'s internally more complex data manipulation functions produce much greater overheads, which can noticeably add-up in longer scripts.\footnote{\fct{bmark} is a slim wrapper around \fct{bench::mark}. See the Computational details section.}

@@

bmark(collapse = collapse::fsummarise(wlddev, mean = fmean(PCGDP)),
      dplyr = dplyr::summarise(wlddev, mean = fmean(PCGDP)))

@@

Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices. For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions like \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 45 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}

@@

Time series functions such as \fct{fgrowth} to compute growth rates are S3 generic and can be applied to most time series classes. In addition to a \code{g} argument for grouping, these functions also have a \code{t} argument for indexation. But first, I provide a basic example of computing the annualized 10-year growth rates in miles flown by airlines in the USA from 1937-1960.

@@

fgrowth(airmiles, n = 10, power = 1/10) |> na.omit() |> round(1)

@@

The results show that the flight volume as been growing steadily but at a decreasing rate. \newline

To illustrate the full capabilities of these time series functions, I generate a sector-level trade dataset of export values (v) by country (c), sector (s), and year (y). Like many detailed trade datasets, it is unbalanced---not all sectors/products are exported by country c in all years.
<<exports>>=
set.seed(101)
exports <- expand.grid(y = 2001:2010, c = paste0("c", 1:10),
  s = paste0("s", 1:10)) |> fmutate(v = abs(rnorm(1e3))) |>
  colorder(c, s) |> fsubset(-sample.int(1e3, 500))
@
The following extracts one country-sector series from the \code{exports} dataset. It is irregular, missing years 2003 and 2006.\footnote{\code{\%=\%} is an infix operator for the \fct{massign} function in \pkg{collapse} which is a multivariate version of \fct{assign}.} Indexation using the \code{t} argument still allows for correct (time-aware) computations in this context without the need to 'expand' the data/fill gaps.
%
<<ts_example_2>>=
.c(y, v) %=% fsubset(exports, c == "c1" & s == "s7", -c, -s)
print(y)
fgrowth(v, t = y) |> round(2)
fgrowth(v, -1:3, t = y) |> head(4)
@
%
If \code{t} is a plain numeric vector or factor as in this case, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a plain numeric vector representing calendar years. Users should manually call \code{timeid()} on plain vectors with decimals.} Time objects like \class{Date} or \class{POSIXct} on the other hand are internally passed through \fct{timeid} to generate an appropriate integer representation of them.\footnote{\fct{timeid} divides by the greatest common divisor (GCD) and subtracts the minimum to generate an integer-id (starting form 1). For this approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}.} \newline

@@

These functions and operators are also integrated with \fct{fgroup\_by} and \fct{fmutate} for vectorized computations. As mentioned earlier, ad-hoc grouping is always more efficient.

@@

to the creation of new classes \class{indexes\_series} and \class{indexed\_frame} which inherit from the former. Any data frame-like object can become an \class{indexed\_frame} and behave as usual for other operations. The technical implementation of these classes is described in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{documentation}. Their basic usage is:

@@

Indexation can be undone using \code{unindex()} and redone with \code{reindex()} and a suitable \class{index\_df}. \class{indexes\_series} can be atomic vectors or matrices (including \class{ts} or \class{xts}) and can also be created directly using \code{reindex()}.

@@

bmark(base = unique(g_int), collapse = funique(g_int))
bmark(base = unique(g_char), collapse = funique(g_char))
